---
title: "Cyclistic Bike Data Analysis Report"
author: "Sayyam K Nahar"
date: "2023-12-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

In 2016, Cyclistic launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that are
geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to
any other station in the system anytime.


### Ask

**Business Question:**

How do annual members and casual riders use Cyclistic bikes differently?

**Key Stackholder:**

* Lily Moreno (Director Of Marketing)
* Cyclistic Marketing Analytics Team
* Cyclistic Executive Team


### Prepare

The data contains information about Cyclistic's historical trip & customer bike usage pattern. The data is obtained from December 2022 to November 2023(12 Months of data).

Data Source: [link](https://divvy-tripdata.s3.amazonaws.com/index.html)


### Process

R Studio will be used as Data Cleaning, Manipulation and Vizualization tool.


**Data Preparation:**

_First, We need to upload lubricate, tidyverse packages._

```{r}
   library(tidyverse)
   library(lubridate)
```

_Set working directory._

```{r}
   setwd("/Users/SAYYAM NAHAR/Desktop/Projects/Case Study for Portfolio/Google DA- CS1 Cyclistic/Dataset")
```

_Import the data from December 2022 to November 2023._

```{r}
   dec_2022 <- read.csv("202212-divvy-tripdata.csv")
   jan_2023 <- read.csv("202301-divvy-tripdata.csv")
   feb_2023 <- read.csv("202302-divvy-tripdata.csv")
   mar_2023 <- read.csv("202303-divvy-tripdata.csv")
   apr_2023 <- read.csv("202304-divvy-tripdata.csv")
   may_2023 <- read.csv("202305-divvy-tripdata.csv")
   jun_2023 <- read.csv("202306-divvy-tripdata.csv")
   jul_2023 <- read.csv("202307-divvy-tripdata.csv")
   aug_2023 <- read.csv("202308-divvy-tripdata.csv")
   sep_2023 <- read.csv("202309-divvy-tripdata.csv")
   oct_2023 <- read.csv("202310-divvy-tripdata.csv")
   nov_2023 <- read.csv("202311-divvy-tripdata.csv")
```

_Check the column names of the imported data._

```{r}
  colnames(dec_2022)
  colnames(jan_2023)
  colnames(feb_2023)
  colnames(mar_2023)
  colnames(apr_2023)
  colnames(may_2023)
  colnames(jun_2023)
  colnames(jul_2023)
  colnames(aug_2023)
  colnames(sep_2023)
  colnames(oct_2023)
  colnames(nov_2023)
```

_Since all the column name are same there is no need to make any changes with it._
_we merge the 12 month of data into single dataframe._

```{r}
all_trips <- bind_rows(dec_2022,jan_2023,feb_2023,mar_2023,apr_2023,may_2023,jun_2023,jul_2023,aug_2023,sep_2023,oct_2023,nov_2023)
```


**Data Cleaning**

_Remove start_lat, start_lng, end_lat and end_lng columns from data._

```{r}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))
```

_Create date, day, month, year, and day of week column for further analysis._

```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

_Create ride_length column using ```difftime()``` function._

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```


_Convert ride_length column into numeric._

```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```


_Remove data that have zero or negative value in ride_length column._

```{r}
all_trips_v2 <- all_trips[!(all_trips$ride_length<0),]
```

_Arrange the data in proper order._

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```


### Analysis

_Calculate mean, median and finding the minimum and maximum value for ride_length column._

```{r}
mean(all_trips_v2$ride_length)
median(all_trips_v2$ride_length)
max(all_trips_v2$ride_length)
min(all_trips_v2$ride_length)
```

_Comparing member and casual user._
_Calculating mean of ride_length for members and casual users using ```aggregate()``` function_

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
```

_Calculating median of ride_length for members and casual users._

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median) 
```

_Calculating max of ride_length for members and casual users._

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
```

_Calculating min of ride_length for members and casual users._

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

_Checking avg ride_length time by each day for member vs casual user._

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

_Analyzing the ridership data by user type and week day._

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
```




### Vizualisation

_1. Visualizing distribution of member and casual riders._

```{r}
all_trips_v2 %>% 
  group_by(member_casual) %>% 
  summarise(cnt = n()) %>%
  mutate(percentage = cnt/sum(cnt)) %>% 
  ggplot(aes(x = "", y = cnt, fill = member_casual)) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar("y", start=0) +
  geom_text(aes(label = scales::percent(percentage)), position = position_stack(vjust = 0.5)) +
  labs(title = 'Distribution of Annual and Casual Riders(in percentage)')+
  theme_void()
```


_2. Visualization for number of rides by rider type and member type._

```{r}
all_trips_v2 %>%
  group_by(member_casual, rideable_type) %>% 
  summarise(cnt2 = n()) %>% 
  ggplot(aes(x = rideable_type, y = cnt2, fill = member_casual)) +
  geom_bar(position = "dodge", stat = "identity", color = "white", width = 0.7) +
  labs(title = "Total Number of Rides by Bike Types and Member Types", x = 'Type of Bikes', y = 'Number of Rides')
```

_3. Visualization for number of rides by member and day of week._

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_bar(position = "dodge", stat = "identity", color = "white", width = 0.7) +
  labs(title = "Total Number Of Rides by Member Type and Week of Day", x = "Weekday", y = "Number of rides")
```

_4. Visualization for number of rides by member and month._

```{r}
all_trips_v2 %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(member_casual, month) %>% 
  ggplot(aes(x=month, y = number_of_rides, fill=member_casual)) +
  geom_bar(position = "dodge", stat = "identity",color = "white", width = 0.7)+
  labs(title = 'Total Number of Rides by Member Types and Months',x = 'Month',y = 'Number of Rides')
```

_5. visualization for average duration._

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_bar(position = "dodge", stat = "identity", color = "white", width = 0.7)+
  labs(title ='Average Duration Of Rides During The Week By Member Type', x = 'Weekday', y = 'Average Duration')

```

